---
title: "Analysis of the Change of Ozone Concentration"
subtitle: "Chapter 5 Report Exercise -- Data Visullization"
author: "Ting Tan"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_align: 'center'
---

```{r setup, include=FALSE, eval = TRUE}
knitr::opts_chunk$set(
  comment = "#>", echo = FALSE, fig.width = 12, message = FALSE
)

library(dplyr)
library(ggplot2)
library(tidyr)  
library(lubridate)
library(cowplot)
```

## Introduction

Ground-level ozone has significant impacts on both the environment and human health. While ozone in the upper atmosphere protects life by absorbing harmful ultraviolet (UV) radiation, elevated concentrations at ground level contribute to air pollution, causing respiratory issues and other health problems. Understanding ozone trends and the factors influencing its formation is essential for effective air quality management and public health protection.

Ozone levels vary over time, typically peaking in warmer months due to increased sunlight and temperature, which enhance photochemical reactions. Key influencing factors include temperature, solar radiation, wind speed, and the presence of pollutants such as nitrogen oxides (NOₓ) and volatile organic compounds (VOCs), especially in urban areas.

## Research Question

* **What are the main factors influencing ground-level ozone concentration? **
* **How does solar radiation, wind and temperature affect the concentration of ground-level ozone? **

## Available data

This analysis uses the airquality dataset in R, which contains 153 daily observations recorded in New York from May to September 1973. The data includes:

* Ozone (ppb) – Measured between 1–3 PM at Roosevelt Island
* Solar Radiation (lang) – Recorded 8–12 AM at Central Park
* Wind Speed (mph) – Morning average at LaGuardia Airport
* Temperature (°F) – Daily maximum at LaGuardia Airport
* Month and Day

###  Data Overview

#### Missing Values

Out of 153 observations, 42 contain missing values in Ozone and Solar.R, mostly occurring in June. These incomplete records are excluded from the analysis to ensure data quality. (See Figure 1a)

#### Initial Trend

Figure 1b shows that ozone concentrations fluctuate notably during midsummer, with sharp, short-term spikes. The missing data in June limits full interpretation of trends during that period.

```{r Data_Cleaning, eval = TRUE}
# Load the airquality dataset
data("airquality")

# Inspect the dataset
# str(airquality)
# summary(airquality)

# Convert Month and Day into a Date format
airquality <- airquality |>
  dplyr::mutate(
    Date = lubridate::ymd(paste(1973, Month, Day, sep = "-"))
  )

# Convert to long format
airquality_long <- airquality |>
  dplyr::select(-Month, -Day) |>
  tidyr::pivot_longer(cols = -Date, names_to = "Variable", values_to = "Value") |>
  dplyr::mutate(IsMissing = ifelse(is.na(Value), "1", "0"))

# Plot missing values heatmap
plot_mis <- ggplot2::ggplot(airquality_long, aes(x = Date, y = Variable, fill = factor(IsMissing))) +
  geom_tile(color = "white", alpha = 0.5) +
  scale_fill_manual(values = c("1" = "indianred2", "0" = "gray80"),
                    name = "Ismissing")+
  labs(title = "Raw data missing values heat map",
       x = "Observation", y = "Variable") +
  theme_classic()

# Remove rows with missing values
airquality_clean <- na.omit(airquality)

# Create the data and check the difference between consecutive dates
airquality_clean <- airquality_clean |>
  arrange(Date) |>
  mutate(
    Date_diff = c(diff(Date), 1),  # Calculate the difference with the next data
    Isgap = ifelse(Date_diff > 1, 1, 0)
  )

# Generate background data for highlighting gaps
background_data <- airquality_clean |>
  mutate(
    xmin = Date,
    xmax = lead(Date, default = max(Date) + 1),  # End date for each rectangle
    # No need for background_color column now
  )

# Plotting
plot_o <- ggplot(airquality_clean, aes(x = Date, y = Ozone)) +
  # Background rectangles to indicate gaps
  geom_rect(data = background_data, 
            aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = as.factor(Isgap)), 
            alpha = 0.5) +  # Map 'Gap' directly to 'fill'
  # Line and points for Ozone concentration
  geom_line(linewidth = 0.5) +
  geom_point(size = 0.8) +
  # Manually add fill scale for the background (legend)
  scale_fill_manual(values = c("0" = "gray80", "1" = "indianred2"), 
                    name = "Data quality", 
                    labels = c("Observed", "Missing")) +
  labs(
    title = "Temporal trend of ozone concentration from cleaned data",
    x = "Date",
    y = "Ozone (ppb)"
  ) +
  theme_classic() +
  theme(legend.position = "right")  # Position the legend to the right
```

```{r ozone_plot, fig.cap = "Figure 1: (a) Missing value distribution of all variables in the airquality dataset; (b) Temporal trend of Ozone concentration after data cleaning (i.e., removing all observations containing missing values).", fig.align = 'center'}

# Display the plot
cowplot::plot_grid(plot_mis, plot_o, ncol = 1, labels = "auto", align = "v")
```

## Temporal Variation Analysis

To explore how ozone levels vary over time, we first compare their trends with those of solar radiation, wind speed, and temperature.

* Temperature shows a similar trend to ozone, suggesting a positive relationship.
* Wind Speed appears negatively related—higher ozone levels tend to occur on calmer days.
* Solar Radiation does not show a clear pattern in relation to ozone levels.

```{r standardized_time_series_plot_separated, fig.cap = "Figure 2: Temporal variations of standardized Ozone, Temperature, Wind Speed, and Solar Radiation, plotted separately.", fig.align = 'center'}
# Standardize the variables: Ozone, Temp, Wind, Solar.R
airquality_standardized <- airquality_clean %>%
  mutate(
    Ozone = scale(Ozone),
    Temp = scale(Temp),
    Wind = scale(Wind),
    Solar.R = scale(Solar.R)
  )

# Create individual plots for each variable
plot_ozone <- ggplot(airquality_standardized, aes(x = Date, y = Ozone)) +
  geom_line(color = "blue") +
  geom_point(size = 1) +
  labs(title = "Standardized Ozone", x = NULL, y = NULL) +
  theme_minimal()

plot_temp <- ggplot(airquality_standardized, aes(x = Date, y = Temp)) +
  geom_line(color = "red") +
  geom_point(size = 1) +
  labs(title = "Standardized Temperature", x = NULL, y = NULL) +
  theme_minimal()

plot_wind <- ggplot(airquality_standardized, aes(x = Date, y = Wind)) +
  geom_line(color = "green") +
  geom_point(size = 1) +
  labs(title = "Standardized Wind Speed", x = NULL, y = NULL) +
  theme_minimal()

plot_solar <- ggplot(airquality_standardized, aes(x = Date, y = Solar.R)) +
  geom_line(color = "purple") +
  geom_point(size = 1) +
  labs(title = "Standardized Solar Radiation", x = NULL, y = NULL) +
  theme_minimal()

# Combine the plots in a grid layout (2 rows and 2 columns)
combined_standardized_plot <- plot_grid(plot_ozone, plot_temp, plot_wind, plot_solar, 
                                        ncol = 2, nrow = 2, labels = "auto", align = "hv")

# Display the combined plot
combined_standardized_plot

```

## Implementation

To further investigate these relationships, we conducted separate linear regression analyses of ozone concentration against each meteorological variable. This approach helps quantify each variable’s influence on ozone levels and visualize the strength and direction of the effects.

```{r MainFuction}
process_models <- function(df_clean){
  # Create an empty list to store each plot
  plot_list <- list()
  
  meta_data <- list(
    variable = c("Solar.R", "Wind", "Temp"),
    formula = c("Ozone ~ Solar.R", "Ozone ~ Wind", "Ozone ~ Temp"),
    lab_txt = c("Solar Radiation (lang)", "Wind Speed (mph)", "Temperature (°F)")
  )
  
  for (i in 1:length(meta_data$variable)) {
    # Dynamically generate the regression formula
    formula <- as.formula(meta_data$formula[i])
    
    # Compute the regression model
    model <- lm(formula, data = df_clean)
    summary_model <- summary(model)
    
    # Extract R² and p-value
    r_squared <- round(summary_model$r.squared, 3)
    p_value <- summary_model$coefficients[2, 4]  # p-value is the 4th element in the second row

    # Format p-value text
    p_text <- if (p_value < 0.005) {
      "P < 0.005"
    } else {
      paste("P =", round(p_value, 3))
    }
    
    # Create annotation text
    annotation_text <- paste("R² =", r_squared, "\n", p_text)
    
    # Plot each variable against Ozone
    suppressWarnings(
      p <- ggplot(df_clean, aes_string(x = meta_data$variable[i], y = "Ozone")) +
        geom_point(color = "black", alpha = 0.8) +
        geom_smooth(
          method = "lm",
          se = FALSE,
          color = "red",
          linewidth = 0.5
        ) +
        annotate(
          "text",
          x = max(df_clean[[meta_data$variable[i]]], na.rm = TRUE) - 5,
          y = max(df_clean$Ozone, na.rm = TRUE) - 10,
          label = annotation_text,
          hjust = 1,
          size = 3,
          color = "black"
        ) +
        labs(x = meta_data$lab_txt[i], y = NULL) +
        theme_classic()
    )
    # Add the plot to the list
    plot_list[[i]] <- p
  }
  
  # Return the list containing all the plots
  return(plot_list)
}
```

## Results and Interpretation

Figure 3 summarizes the regression results:

* Temperature has a strong positive correlation with ozone. Warmer days are linked to higher ozone levels, consistent with photochemical ozone formation.
* Wind Speed shows a clear negative correlation. Stronger winds help disperse pollutants, leading to lower ozone concentrations.
* Solar Radiation has a weaker and less consistent relationship with ozone, suggesting a more complex or indirect influence.

These findings highlight the roles of temperature and wind in shaping urban ozone pollution. Such insights are valuable for informing air quality policies, especially in cities where temperature and wind patterns can exacerbate pollution episodes.

```{r regression_plot, fig.cap = "Figure 3: Regression analysis of Ozone concentration with environmental variables. (a) shows Solar Radiation, (b) shows Wind Speed, and (c) shows Temperature. Each plot includes the regression line with R² and p-value annotations.", fig.align = 'center'}
# Generate the individual plots
plots <- process_models(airquality_clean)

combined_plot <- plot_grid(plotlist = plots, ncol = 3, labels = c("a", "b", "c"), align = "hv")

final_plot <- ggdraw() +
  draw_label("Ozone Concentration (ppb)", x = 0.02, y = 0.5, angle = 90, vjust = 1, size = 12) +
  draw_plot(combined_plot, x = 0.05, y = 0, width = 0.95, height = 1)

final_plot