---
title: "Analyzing changes in soil organic matter during elevated CO₂ experiments"
subtitle: "Chapter 4 Report Excercise -- Data Wrangling"
author: "Ting Tan (22-108-534)"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2
bibliography: literatures.bib
csl: apa.csl
---

```{r setup, include=FALSE, eval = TRUE}
knitr::opts_chunk$set(
  comment = "#>",    # Prefixes for printed output from R
  echo = FALSE,      # Hides R code in the final document
  fig.width = 6,     # Sets width of figures (in inches)
  message = FALSE,
  warning = FALSE
)

library(dplyr)
library(tidyr)
library(readr)
library(knitr)
library(ggplot2)
library(kableExtra)
```

---

# Introduction

The $CO_2$ fertilization effect refers to the enhancement of plant growth resulting from elevated atmospheric carbon dioxide ($CO_2$) concentrations. As $CO_2$ is a fundamental substrate for photosynthesis, higher levels can stimulate photosynthetic rates in many plant species, leading to increased biomass production. This enhanced plant productivity can result in greater inputs of soil organic carbon ($SOC$) through leaf litter, root exudates, and other plant residues. However, increased carbon inputs may also stimulate microbial activity, potentially accelerating the decomposition of existing soil organic carbon—a process known as priming. Thus, while elevated $CO_2$ can initially promote carbon sequestration through increased plant growth, its long-term impact on soil carbon storage is complex and may be counteracted by faster microbial decomposition rates [@groenigen2014].

---

## Research Question and Hypothesis

**How does the decomposition rate of soil organic carbon change response to the increased atmospheric $CO_2$ concentration?**

**Hypothesis**: Faster decomposition under increased atmospheric $CO_2$  limits soil carbon storage [@groenigen2014].

---

## Available Data

This project uses *Database S1*, which was downloaded from the supplementary materials of the paper by @groenigen2014. *Database S1* contains data of soil organic carbon measurements in experiments, where ecosystems  are exposed to ambient (low) and elevated (high) $CO_2$ concentrations ("$mean$", $g C m_2$). The mean soil organic carbon  of multiple samples (“$n$”, $g C m_2$) is recorded within each experiment for different sample dates. Information  is provided for the time in years since the start of the experiment (“Time (years)”). We manually clean the data in the tab “Database S1” and save it  as a CSV file that can be read into R in the the project data folder (`./data_raw/1249534s1.csv`)

---

# Implementation

To evaluate the effect of elevated atmospheric CO₂ on soil organic carbon (SOC), we performed the following steps in R:

1. **Data Preparation:** Imported and cleaned the dataset, handled missing values, and renamed columns.
2. **Log-Response Ratio (LRR):** Calculated LRR for CO₂ concentration and SOC using
   `LRR = ln(x_elevated / x_ambient)` for each paired observation within experiments.
3. **Phase Classification:** Categorized observations into three temporal phases:

   * Early (<3 years), Mid (3–6 years), Late (>6 years).
4. **Aggregation:** Computed the mean LRR and standard error for each phase, based on individual LRRs rather than treatment means, to assess how CO₂ effects on SOC evolve over time.

```{r DataClean, eval = TRUE}
# Read the soil organic carbon measurement dataset, skipping meta-data
soc_values <- readr::read_csv(here::here("data_raw/1249534s1.csv"), skip = 3)

# select variables for analysis
soc_values_clean <- soc_values |>
  select(`Experiment`,
         `Time (years)`,
         starts_with("ambient"), 
         starts_with("increased"))

# Remove empty rows
soc_values_clean <- soc_values_clean[rowSums(!is.na(soc_values)) > 0, ]

# Remove empty columns
soc_values_clean <- soc_values_clean[, colSums(!is.na(soc_values_clean)) > 0]

# Fill missing values
soc_values_clean <- soc_values_clean |> 
  tidyr::fill(everything(), .direction = "down")

# Rename columns for better readability
soc_values_clean <- soc_values_clean |>
  dplyr::rename(
    CO2_elevated = `increased CO2...7`, 
    CO2_ambient = `ambient CO2...6`,      
    SOC_elevated = `increased CO2...9`,   
    SOC_ambient = `ambient CO2...8`,
    year = `Time (years)`
  ) 
```


```{r LRR, eval = TRUE}
# Data calculations: this part will include log-transformed response ratios (LRR) for CO2 and SOC
soc_co2_lrr <- soc_values_clean |>

  # Calculate log-response ratios for each experiment (log of elevated / ambient)
  dplyr::mutate(
    CO2_lrr_exp = log(CO2_elevated / CO2_ambient),  # Log-transformed ratio for CO2
    SOC_lrr_exp = log(SOC_elevated / SOC_ambient)   # Log-transformed ratio for SOC
  ) |>
  
  # Categorize data into experimental phases based on time (years)
  dplyr::mutate(phase = case_when(
      `year` < 3 ~ "Early",   # Phase 1: Early (less than 3 years)
      `year` >= 3 & `year` <= 6 ~ "Mid",   # Phase 2: Mid (3 to 6 years)
      `year` > 6 ~ "Late"   # Phase 3: Late (more than 6 years)
  )) |>
  
  # Calculate mean log-response ratio for each phase by aggregating the data
  group_by(phase) |>
  mutate(
    CO2_lrr_pha = mean(CO2_lrr_exp),  # Mean log ratio for CO2 per phase
    SOC_lrr_pha = mean(SOC_lrr_exp)  # Mean log ratio for SOC per phase
    # CO2_lrr_pha = log(mean(CO2_elevated, na.rm = TRUE) / mean(CO2_ambient, na.rm = TRUE)),  # Mean log ratio for CO2 per phase
    # SOC_lrr_pha = log(mean(SOC_elevated, na.rm = TRUE) / mean(SOC_ambient, na.rm = TRUE))   # Mean log ratio for SOC per phase
  ) |> 
  
  # Ungroup the data after calculation to remove the grouping structure
  ungroup()

# Save the cleaned and calculated data as an RDS file
write_rds(soc_values_clean, 
          file = here::here("data", "soc_co2_lrr.rds"))


```
---

# Results and Discussion

From Table \@ref(tab:phaselrr), we observe that the SOC response generally follows the same direction as changes in atmospheric CO₂ concentration. During the early phase of exposure, SOC increases at a faster rate than atmospheric CO₂, suggesting a strong initial accumulation of organic carbon in soil. However, this accumulation slows down in the mid phase and eventually reverses in the late phase, with SOC levels declining over time. This pattern—also illustrated in Figure \@ref(fig:explrr)—indicates that although elevated CO₂ may stimulate carbon input initially, it also promotes faster decomposition, which ultimately limits SOC accumulation.

```{r phaselrr, results='asis'}

soc_co2_lrr |>
  select(phase, CO2_lrr_pha, SOC_lrr_pha) |>  
  distinct() |>
  mutate(
    CO2_lrr_pha = round(CO2_lrr_pha, 4),
    SOC_lrr_pha = round(SOC_lrr_pha, 4)
  ) |>
  kable(
    col.names = c("Phase", "CO₂ LRR (mean of phase)", "SOC LRR (mean of phase)"),
    caption = "Summary of Log-Response Ratios by Phase",
    format = "html",
    booktabs = TRUE
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "center"
  ) |>
  column_spec(1, width = "10em") |>
  column_spec(2, width = "15em") |>
  column_spec(3, width = "15em")
```

```{r explrr, fig.align='center', fig.height=3, fig.cap="Scatterplot showing the relationship between the log-response ratio (LRR) of atmospheric CO₂ concentration and the LRR of soil organic carbon (SOC), including a linear regression trend line."}
ggplot(soc_co2_lrr, aes(x = CO2_lrr_exp, y = SOC_lrr_exp, colour = year)) +
  geom_point(alpha = 0.5, size = 2.5) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed") +
  scale_colour_gradientn(colours = c("#FF6F91", "#FF9671", "#FFC75F", "#F9F871", "#82CD47", "#47B5FF", "#845EC2")) +
  labs(
    x = expression("LRR of " * CO[2] * " concentration"),
    y = expression("LRR of soil organic carbon"),
    colour = "year"
  ) +
  theme_classic()

```

---

# Reference
```{r, include=FALSE, eval=FALSE}
# This code will neither run nor show in the output
rmarkdown::render(here::here("vignettes/re_tidy.Rmd"))
```


